The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-09-29 thesis/extra_rv_models.tex
thesis edits

\chapter{Radial velocity modeling}\label{sec:rv_obs}

@@

For both Sculptor and Ursa Minor, we construct literature samples of
radial velocity measurements. We combine these samples with J+24's
members to produce RV consistent stars and to compute velocity
dispersion, systematic velocities, and test for the appearance of
velocity gradients.

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1-3
arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
mean RV measurement from each study using the inverse-variance weighted
mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\). We remove stars with significant velocity dispersions
as measured between observations in a study or between studies. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically).

Next, we need to correct the coordinate frames for the solar motion and
on-sky size of the galaxy. We transform the frame into the galactic
standard of rest (GSR). The next step is to account for the slight
differences in the direction of each radial velocity. Let the \(\hat z\)
be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
the angular distance between the centre of the galaxy and the individual
star, the corrected radial velocity is then \begin{equation}{
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
}\end{equation} where \(v_{\rm los, gsr}\) is the line of sight velocity
in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the tangental
velocities in RA and Dec, and \(\theta\) is the position angle of the
star with respect to the centre of the dwarf. The correction from both
effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We then use the \(v_z\) values for the
following modelling, however repeating with uncorrected, heliocentric
velocities does not significantly affect the results.

The combined likelihood, including RV information, becomes
\begin{equation}{
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
}\end{equation} where we assume that the satellite and background
distributions are Gaussian. Specifically, \begin{equation}{
\begin{split}
{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
}\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
will dominate the uncertainty budget here. We assume a halo/background
velocity dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.

Similar to above, we retain stars with the resulting membership
probability of greater than 0.2. Because of the additional information
from radial velocities, most stars have probabilities close to 1 or 0 so
the probability cut is not too significant.

We assume priors on systematic velocity and velocity dispersion of
\begin{equation}{
\begin{split}
\mu_{v} &= N(0\,\kms, \sigma_{\rm halo}^2) \\ 
\sigma_{v} &= U(0, 20\,\kms)
\end{split}
}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
the velocity dispersion of the MW halo adopted above, a reasonable
assumption for dwarfs in orbit around the MW.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[LOS velocity fit to Scl.]{Velocity histogram of Scl and UMi in
terms of \(v_z\) (REF). Orange points are from our crossmatched RV
membership sample.}
\end{figure}

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}. See appendix REF for a more
detailed comparison between different samples and additional tests.

We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
a position angle of \(-149_{-13}^{+17}\) degrees (see appendix REF).
Several past work has attempted to detect a gradient in Sculptor, but no
consensus has been reached. \citet{arroyo-polonio+2024} detect a
velocity gradient of \(4\pm1.5\,\kmsdeg\) in a similar direction using
the \citet{tolstoy+2023} sample, finding inconclusive statistical
evidence. They additionally suggest a third chemodynamical component of
the galaxy which may bias rotation measurements. \citet{battaglia+2008}
also detect a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along
the major axis, approximately the same direction. Instead,
\citet{strigari2010}; \citet{martinez-garcia+2023} detect no significant
gradient in Sculptor using \citet{WMO2009} sample. Note that
pre-\emph{Gaia} work did not have as strong of a constraint on the
proper motion of Scl, which limits conclusions of the intrinsic velocity
gradient in Scl.

\begin{itemize}
\tightlist
\item
  Inter-study systematics and biases. While basic crossmatches and a
  simple velocity shift, combining data from multiple instruments is
  challenging. This appears to be a minor issue (Sculptor) or is
  corrected for (Ursa Minor).
\item
  Misrepresentative uncertainties. Inspection of the variances compared
  to the standard deviations within a study seems to imply that errors
  are accurately reported. APOGEE notes that their RV uncertainties are
  known to be underestimates but are a small proportion of our sample.
\item
  Binarity. While not too large of a change for classical dwarfs, this
  could inflate velocity dispersions of about \(9\,\kms\) by about
  \(1\,\kms\)\citep{spencer+2017}. Thus, our measurement is likely
  slightly inflated given the high binarity fractions measured in these
  systems \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
\item
  Multiple populations. Both Sculptor and Ursa Minor likely contain
  multiple populations \citep[\citet{pace+2020},
  \citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a
  single population, and each population may have a different extent and
  velocity dispersion, this could result in biased velocity dispersions.
  However, it is unclear how to uniquely determine an overall velocity
  dispersion in a multi-population system.
\item
  Selection effects. RV studies each have their own selection effects,
  which may affect the resulting dispersion, especially if different
  populations or regions of the galaxy have different velocities or
  velocity dispersions. We do not attempt to correct for this.
\end{itemize}

For both Ursa Minor and Sculptor, we also fit models to only data from
individual surveys (see REF). Since the resulting parameters are very
similar, we conclude that many of the systematic uncertainties are
likely smaller than the present errors or that each large survey has
similar biases.

\citet{martinez-garcia+2023}: rotation gradients.

\section{Velocity modelling and
comparisons}\label{velocity-modelling-and-comparisons}

Here, we describe in additional detail, our methods and comparisons for
RV modelling between studies.

Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
smoothed samples from posterior/prior.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
\caption[Scl velocity sample]{RV members of Sculptor plotted in the
tangent plane coloured by corrected velocity difference from mean
\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
scale).}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
\caption[Scl velocity gradient]{The corrected LOS velocity along the
best fit rotational axis. RV members are black points, the systematic
\(v_z\) is the horizontal grey line, blue lines represent the
(projected) gradient from MCMC samples, and the orange line is a rolling
median (with a window size of 50).}
\end{figure}

@@

>>>>

\chapter{LOS velocity modeling}\label{sec:rv_obs}

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Scl and UMi. Our goal is to derive the systemic
velocity and velocity disperions, and understand if the galaxies show
evidence of tidal features, e.g.~a velocity gradient or a rising outer
velocity dispersion. Our derived velocity distributions are consistent
with literature. We find weak evidence for a velocity gradient in Scl
and a rising outer velocity dispersion. However, Scl's velocity gradient
is misaligned with its proper motion, likely inconsistent with a tidal
interpretation. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
observable features of tidal disruption given current velocity
observations.

@@

For both Sculptor and Ursa Minor, we construct literature samples of
radial velocity measurements. We combine these samples with J+24's
Bayesian likelihoods to produce likely members using additional RV
information.

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1--3
arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
mean RV measurement from each study using the inverse-variance weighted
mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\).

We remove stars with significant velocity dispersions as measured
between observations in a study or between studies, to filter out
unreliable measurements or possible spectroscopic binaries. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically)..

Next, we need to correct the coordinate frames for the solar motion and
on-sky size of the galaxy. We transform the frame into the galactic
standard of rest (GSR). The next step is to account for the slight
differences in the direction of each radial velocity. Let the \(\hat z\)
be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
the angular distance between the centre of the galaxy and the individual
star, the corrected radial velocity is then \begin{equation}{
v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the line of
sight velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
tangental velocities in RA and Dec, and \(\theta\) is the position angle
of the star with respect to the centre of the dwarf. The correction from
both effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We then use the \(v_z\) values for the
following modelling, however repeating with uncorrected, heliocentric
velocities does not significantly affect the results.

The combined likelihood, including RV information, becomes
\begin{equation}{
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
}\end{equation} where we assume that the satellite and background
distributions are Gaussian. Specifically, \begin{equation}{
\begin{split}
{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
}\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
dominates the uncertainty budget. We assume a halo/background velocity
dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.

Similar to above, we retain stars with the resulting membership
probability of greater than 0.2. Like in J+24, most stars have
probabilities of nearly 1 or 0, so the precise choice of cut marginally
affects the resulting sample.

\section{MCMC modeling}\label{mcmc-modeling}

We use Monte-Carlo Markov chains (MCMC) to calculate the posterior
distributions of our models of the intrinsic velocity dispersion and
possible velocity gradients in Scl and UMi.

The likelihood is based on the distribution of stars with respect to the
model's predicted \(\mu_\V\) and \(\sigma_\V\) at each star's position
in the tangent plane \(\xi\) and \(\eta\). Specifically, the log
likelihood is \begin{equation}{
\log {\cal L} = \sum_i \frac{1}{\sigma_i} \log f\left(\frac{\V_i - \mu_i}{\sigma_i}\right),
}\end{equation} where \(f\) is the probability distribution of a
standard normal distribution, \(\V_i\) is the velocity of the \(i\)th
star, and \(\mu_i\) and \(\sigma_i\) are the model's predicted mean and
velocity dispersion at the \(i\)th star's position \(\xi_i\) and
\(\eta_i\).

For all models, we ado \begin{equation}{
\mu_{0} \sim N(0\,\kms, \sigma^2_{\rm halo})\\
\sigma_{0} \sim U(0\,\kms, 20\,\kms),
}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
the velocity dispersion of the MW halo adopted above, a reasonable
assumption for dwarfs in orbit around the MW.

For the simplest model, \(\mu\) and \(\sigma\) are held
fixed---\(\mu(\xi, \eta) = \mu_0\) and \(\sigma(\xi, \eta) = \sigma_0\).
There are no additional parameters besides \(\mu_0\) and \(\sigma_0\).

To model a velocity gradient, we add two parameters, the gradient
\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta\\
\sigma = \sigma_0
}\end{equation}

To model an increasing velocity dispersion, we add a third parameter,
\(c\) which applies as \begin{equation}{
\log \sigma(R_{\rm ell}) = \log \sigma_0 + c\,\log\left(R_{\rm ell} / R_h\right) \\
\mu = \mu_0 + a\,\xi + b\,\eta\\
}\end{equation} with \(c \sim U(0, 1)\) where \(R_{\rm ell}\) is the
elliptical radius. We only consider this model including the velocity
gradient as a velocity gradient is the simplest way to increase the
dispersion gradient.

Bayes factors quantify the relative evidence between Bayesian model. If
the Bayes factor

Our models are \emph{nested}, where the fiducial model (\(\sigma_0\) and
\(\mu_0\) only) always arises from setting parameters of the more
complex models to zero (\(a=0, b=0\) and/or \(c=0\) ). As a result, we
can use the Savage-Dickey method to calculate the Bayes factors
\citep{dickey+lientz1970}, where the Bayes factor is then the relative
density of posterior versus prior samples where \(a=b=c=0\). To
calculate densities, we use a Silverman-bandwidth KDE to smooth the
samples.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[LOS velocity fit to Scl]{Velocity histogram of Scl and UMi in
terms of projected-correction GSR LOS velocity (Eq.~\ref{eq:v_z}). Black
points with error bars are from the crossmatched observed sample, and
green lines represent MCMC fits to the velocity dispersion.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
\caption[Scl velocity gradient]{The corrected LOS velocity along the
best fit rotational axis. RV members are black points, the systematic
\(v_z\) is the horizontal grey line, blue lines represent the
(projected) gradient from MCMC samples, and the orange line is a rolling
median (with a window size of
50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
\caption[Sculptor's velocity gradient orientation]{RV members of
Sculptor plotted in the tangent plane coloured by corrected velocity
difference from mean \(v_z - \bar v_z\) . The black ellipse marks the
half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
arrows mark the proper motion (PM, GSR frame) and derived velocity
gradient (rot) vectors (to scale).}
\end{figure}

@@

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}.

We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
a position angle of \(-149_{-13}^{+17}\) degrees. Several past work has
attempted to detect a gradient in Sculptor, but no consensus has been
reached. \citet{arroyo-polonio+2024} detect a velocity gradient of
\(4\pm1.5\,\kmsdeg\) in a similar direction using the
\citet{tolstoy+2023} sample, finding inconclusive statistical evidence.
They additionally suggest a third chemodynamical component of the galaxy
which may bias rotation measurements. \citet{battaglia+2008} also detect
a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along the major
axis, approximately the same direction. Instead, \citet{strigari2010};
\citet{martinez-garcia+2023} detect no significant gradient in Sculptor
using \citet{WMO2009} sample. Note that pre-\emph{Gaia} work did not
have as strong of a constraint on the proper motion of Scl, which limits
conclusions of the intrinsic velocity gradient in Scl.

Fig.~\ref{fig:scl_velocity_gradient_scatter} plots the combined velocity
sample and the MCMC samples for the velocity gradient. While some
samples have a consistent gradient with 0, most samples have a positive
velocity gradient, consistent with the rolling median trend.

The velocity gradient in Scl is misaligned with the proper motion

For both Ursa Minor and Sculptor, we also fit models to only data from
individual surveys. Since the resulting parameters are very similar, we
conclude that many of the systematic uncertainties are likely smaller
than the present errors or that each large survey has similar biases.

\emph{Inter-study systematics and biases}. While basic crossmatches and
a simple velocity shift, combining data from multiple instruments is
challenging. This appears to be a minor issue (Sculptor) or is corrected
for (Ursa Minor).

\emph{Misrepresentative uncertainties}. Inspection of the variances
compared to the standard deviations within a study seems to imply that
errors are accurately reported. APOGEE notes that their RV uncertainties
are known to be underestimates but are a small proportion of our sample.
We find that most samples report internally-consistent uncertainties for
multiple observations of the same star.

\emph{Binarity}. Stars in binary systems can inflate the inferred
system's velocity dispersion. For example, a system with an apparent
velocity dispersion of \(9\,\kms\) may have a true velocity dispersion
of \(8\,\kms\) \citep{spencer+2017}. Thus, our measurement is likely
slightly inflated given the high binary fractions measured in these
systems \citep{spencer+2018, arroyo-polonio+2023, gration+2025}.

\emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
multiple populations \citep[\citet{pace+2020},
\citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a single
population, and each population may have a different extent and velocity
dispersion, this could result in biased velocity dispersions. However,
it is unclear how to uniquely determine an overall velocity dispersion
in a multi-population system.

\emph{Selection effects}. RV studies each have their own selection
effects, which may affect the resulting dispersion, especially if
different populations or regions of the galaxy have different velocities
or velocity dispersions. We do not attempt to correct for this. Given
that samples extend through the half-light radius of each system, we do
not expect current samples to miss large features except in the very
outskirts.

Could we observe velocity gradients instead in \emph{Gaia}.
Unfortunately, given Scl and UMi's distances (\(\gtrsim 70\,\kpc\))
\emph{Gaia's} systemic proper motion uncertainty of \(0.017\,\masyr\)
corresponds to a velocity of \(5\,\kms\), making the detection of
gradients of amplitude less than \(5\,\kms\) challenging. In addition,
the typical uncertainties of member stars of Scl and UMi are even
larger. While in future data releases, \emph{Gaia} uncertainties may
permit study in the internal motion of dwarfs, the current \emph{Gaia}
proper motions are not yet precise enough.

\section{Summary}\label{summary}

In this section, we analyze literature samples of LOS velocity
measurements for both Scl and UMi. In each case, we find systemic LOS
velocities and velocity dispersions consistent with past work. We detect
weak evidence for a velocity gradient in Scl, however the gradient is
mis-aligned with the proper motion of Scl so is unlikely to be of tidal
origin. In both galaxies, the velocity dispersion profile is consistent
with a constant velocity dispersion with radius. The lack of
observational evidence for ongoing tidal disruption in the velocity
distribution of stars in Scl and UMi further supports our
interpretations in the main text.
